---
format: html
date: "17.05.2024"
university: "University of Lausanne, HEC"
course: "Machine Learning in Business Analytics"
group: "Group E"
year: "2024"
output:
  pdf_document:
    includes:
      in_header: header.tex
geometry: margin=2cm
fontsize: 12pt
header-includes:
  - \usepackage{graphicx}
  - \newcommand{\coverTitle}[1]{\Huge\textbf{#1}\normalsize}
  - \newcommand{\coverAuthor}[1]{\large\textit{#1}\normalsize}
  - \newcommand{\coverInfo}[1]{\small\textbf{#1}\normalsize}
  - \newcommand{\coverDate}[1]{\small{#1}\normalsize}
---

\begin{titlepage}
  \centering
  \vspace*{1cm}
  \includegraphics[width=5cm]{image/Logo_HEC_Lausanne.png}
  \vspace*{1cm}

  \coverTitle{Modeling passenger preferences for air travel upgrades}
  \vspace{1.5cm}
  
  \coverAuthor{Michel Robin, Bickerstaffe Faber and Magnin Antoine}
  \vspace{1.5cm}
  
  \coverInfo{University of Lausanne, HEC}
  \vspace{0.5cm}
  
  \coverInfo{Machine Learning in Business Analytics}
  \vspace{0.5cm}
  
  \coverInfo{2024}
  \vspace{0.5cm}
  
  \coverInfo{Group E}
  \vspace{1cm}
  
  \coverDate{04.05.2024}
  \vfill
\end{titlepage}

\newpage

Abstract: |
 The following machine learning project focuses on...

```{r, echo = FALSE, message = FALSE}
# loading all the necessary packages
source(here::here("src/setup.R"))
```

```{r child = paste0('sections/', c('intro.qmd', 'data.qmd', 'eda.qmd', 'unsupervised.qmd', 'supervised.qmd',  'concl.qmd','references.qmd'))}
```
